---
title: "Divvy Case Study"
author: "Rungchat Amnuay"
date: "2023-03-31"
output:
  html_document: default
  pdf_document: default
mainfont: Time New Roman
papersize: a4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## **Senario**
I am a junior data analyst working in the marketing analyst team at Cyclistic, a bike-share company in Chicago. The director of marketing believes the company’s future success depends on maximizing the number of annual memberships. Therefore, my team wants to understand how casual riders and annual members use Cyclistic bikes differently. From these insights, my team will design a new marketing strategy to convert casual riders into annual members. But first, Cyclistic executives must approve my recommendations, so they must be backed up with compelling data insights and professional data visualizations.

I was assigned to analyze the descriptive insight from the Divvy customer and comparing between casual member and annual member. The business question is "How to convert a casual member to become a annual member.

## **Preparing the data** 
### Step 1: Getting data

I use the public data set from https://divvy-tripdata.s3.amazonaws.com/index.html with 4 data sets to analyze including

  + Divvy_Trips_2019_Q2.csv
  + Divvy_Trips_2019_Q3.csv
  + Divvy_Trips_2019_Q4.csv
  + Divvy_Trips_2020_Q1.csv

All of these files were collected by Divvy itself. Therefore It is Reliable, Original, and Comprehensive

### Step 2: Install packages

  + install.packages("tidyverse")
  + install.packages("dplyr")
  + install.packages("patchwork")

### Step 3: Loading library

```{r load libraly, warning=FALSE, message=FALSE}
library(tidyverse)
library(dplyr)
library(lubridate)
library(hms)
library(ggplot2)
library(patchwork)
library(forcats)
library(scales)
```

### Step 4: Import data

I decided to use RStudio desktop version for a data manipulation tool.
The reasons are

  1. Each file contain a numerous records
  2. RStudio desktop version works on local memory. Then it can work faster than RStudio cloud.

```{r import data}
q2_2019 <- read_csv("Divvy_Trips_2019_Q2.csv")
q3_2019 <- read_csv("Divvy_Trips_2019_Q3.csv")
q4_2019 <- read_csv("Divvy_Trips_2019_Q4.csv")
q1_2020 <- read_csv("Divvy_Trips_2020_Q1.csv")
```
### Step 5: Preview data
Preview the consistency of all files column name and planing for manipulate them before integrate them.

```{r preview data}
colnames(q3_2019)
colnames(q4_2019)
colnames(q2_2019)
colnames(q1_2020)
```
### Step 6: Manipulate data and merge to a single file.
#### **6.1 Renaming column name to be consistence.**


```{r manipulate data}
(q4_2019 <- rename(q4_2019,
                   ride_id = trip_id,
                   rideable_type = bikeid,
                   started_at = start_time,  
                   ended_at = end_time,  
                   start_station_name = from_station_name, 
                   start_station_id = from_station_id, 
                   end_station_name = to_station_name, 
                   end_station_id = to_station_id, 
                   member_casual = usertype))

(q3_2019 <- rename(q3_2019,
                   ride_id = trip_id,
                   rideable_type = bikeid, 
                   started_at = start_time,  
                   ended_at = end_time, 
                   start_station_name = from_station_name, 
                   start_station_id = from_station_id, 
                   end_station_name = to_station_name, 
                   end_station_id = to_station_id, 
                   member_casual = usertype))

(q2_2019 <- rename(q2_2019,
                   ride_id = "01 - Rental Details Rental ID",
                   rideable_type = "01 - Rental Details Bike ID", 
                   started_at = "01 - Rental Details Local Start Time",  
                   ended_at = "01 - Rental Details Local End Time",  
                   start_station_name = "03 - Rental Start Station Name", 
                   start_station_id = "03 - Rental Start Station ID",
                   end_station_name = "02 - Rental End Station Name", 
                   end_station_id = "02 - Rental End Station ID",
                   member_casual = "User Type"))
```

#### **6.2 Covert ride_id and rideable_type to character to stack correctly.**

```{r Convert datatype1}
q4_2019 <-  mutate(q4_2019, ride_id = as.character(ride_id),
                   rideable_type = as.character(rideable_type)) 
q3_2019 <-  mutate(q3_2019, ride_id = as.character(ride_id),
                   rideable_type = as.character(rideable_type)) 
q2_2019 <-  mutate(q2_2019, ride_id = as.character(ride_id),
                   rideable_type = as.character(rideable_type)) 
```

#### **6.3 Merging from 4 files to a single file and drop unrelated columns out of dataframe.**

```{r merge and drop}
all_trips <- bind_rows(q2_2019, q3_2019, q4_2019, q1_2020)

all_trips <- all_trips %>%  
  select(-c(start_lat, 
            start_lng, 
            end_lat, 
            end_lng, 
            birthyear, 
            gender, 
            "01 - Rental Details Duration In Seconds Uncapped", 
            "05 - Member Details Member Birthday Year", 
            "Member Gender", "tripduration"))
tibble(all_trips)
```

### Step 7: Start working with a single dataframe.

#### **7.1 Reinspect the dataframe.**

I started focus on member type based on main objective of analysis. I inspected that column by
```{r inpect member type}
table(all_trips$member_casual)
```
In the member_casual columns, there are something not consistence.

* there are 2 names for annual member: "member" and "Subscriber"
* there are 2 names for casual rider: "Customer" and "casual"

I fixed by
```{r fix the member type}
all_trips <-  all_trips %>% 
  mutate(member_casual = recode(member_casual,
                                "Subscriber" = "member",
                                "Customer" = "casual"))
```
Double-check the result.
```{r reinpect member type}
table(all_trips$member_casual)
```

#### **7.2 Manipulate and create aggregate ride data.**

##### Create Day, Month, Year and Day of Week coloumns for further analysis.
Because in the started_at and ended_at columns contain the YYYY-MM-DD with "POSIXct" and "POSIXt" classes.

```{r create DD,MM,YYYY,DOF columns}
all_trips$date <- as.Date(all_trips$started_at) #The default format is yyyy-mm-dd
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
```

##### Calculate the ride length in a new columns.
```{r calculate ride length}
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)

all_trips$ride_length <- as.numeric(all_trips$ride_length)
```

##### Drop BAD data from data frame and create the final version of dataframe.
In the dataframe include unintentional errors. I decided to ask the project manager for advice and they recommended to drop all of the errors.
```{r drop bad data}
all_trips_v2 <- all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$ride_length<0),]
```

#### Convert Month/Year to Quarter 
```{r convert quarter}
all_trips_v2 <- mutate(all_trips_v2, quarter = quarter(all_trips_v2$started_at)) |>
                  mutate(Q = case_when(
                    quarter == 1 ~ "Q1_2020",
                    quarter == 2 ~ "Q2_2019",
                    quarter == 3 ~ "Q3_2019",
                    quarter == 4 ~ "Q4_2019",
                    TRUE ~ "Unidentified"
                  )) |>
                  mutate(Q = as.factor(Q))
```



## **Analyze the data**

### Descriptive statistics
```{r basic stats}
all_trips_v2 |>
  summarise(total_mean = mean(all_trips_v2$ride_length),
            total_median = median(all_trips_v2$ride_length),
            total_min = min(all_trips_v2$ride_length),
            total_max = max(all_trips_v2$ride_length))

aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)

```
### Member Type Proportion.
```{r member type proportion}
all_trips_v2|>
  group_by(member_casual) |>
  summarise(member_type_n = n())
```

### The average ride length between annual member(member) and casual ride(casual) by day of week.
```{r groupby member and DOF}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + 
            all_trips_v2$day_of_week, FUN = mean)

##reorder day of week
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, 
                                    levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))

aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + 
            all_trips_v2$day_of_week, FUN = mean)
```

### Ride Distribution by Number of Rides and Average Duration Grouped by Member Types and Day of Week
```{r rides dist DOF.}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%  #creates weekday field using wday()
  group_by(member_casual, weekday) %>%  #groups by user type and weekday
  summarise(number_of_rides = n()		#calculates the number of rides and average duration 
            ,average_duration = mean(ride_length)) %>% 	# calculates the average duration
  arrange(member_casual, weekday)	# sorts
```

### The Distribution of Rides by Quarter and Member Type.
```{r quarter distribution}
all_trips_v2 |>
  group_by(Q, member_casual) |>
  summarise(number_of_rides = n())|>
  arrange(Q) |>
  mutate(Q = factor(
    Q,
    levels = c("Q2_2019", "Q3_2019", "Q4_2019", "Q1_2020"),
    ordered = TRUE))
```

<br>

## **Data Visualization** 

### Member Types Proportion
```{r member type proportion graph}
all_trips_v2|>
  group_by(member_casual) |>
  summarise(member_type_n = n()) |>
  ggplot(aes(x = member_casual, y = member_type_n, 
             fill = member_casual)) + 
  geom_col(width = 0.6) + 
  scale_y_continuous(labels = comma) + # fix the y-axis scale 3e+06 bla bla
  labs(
    title = "Types of Rider Distribution",
    x = "Member Type",
    y = "Number of rides",
    caption = "Source: https://divvy-tripdata.s3.amazonaws.com/index.html",
    fill = "Member Type"
  ) +
  geom_text(aes(label = member_type_n),
            position = position_stack(vjust = 1.1)) +
  theme(plot.title = element_text(size = 18))
```

### Number of Rides by Rider Types on Day of Week.
```{r N of rides~type}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") + 
  scale_y_continuous(labels = comma) +
    labs(
    title = "Number of Rides by Rider Types on Day of Week.",
    x = "Day of Week",
    y = "Number of Rides",
    caption = "Source: https://divvy-tripdata.s3.amazonaws.com/index.html",
    fill = "Member Type"
  ) +
  theme(plot.title = element_text(size = 18))
```

### Average Duration between Member Types and Day off Week
```{r avg duration member~DOF}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge") + 
  labs(
    title = "Average Duration by Member Types and Day of Week",
    x = "Day of Week",
    y = "Average Duration (second)",
    caption = "Source: https://divvy-tripdata.s3.amazonaws.com/index.html",
    fill = "Member Type"
  )+
  theme(plot.title = element_text(size = 18))
```

### Top 10 station with highest number of rides stack by member types
```{r top 10 station}
all_trips_v2 |>
  group_by(start_station_name) |>
  summarise(total_ride = n()) |>
  inner_join(
    all_trips_v2 |>
      group_by(start_station_name, member_casual) |>
      summarise(number_of_rides = n()) |>
      arrange(start_station_name, desc(number_of_rides)), by = "start_station_name") |>
  arrange(desc(total_ride), start_station_name) |>
  slice(1:20) |>
  ggplot(aes(x = reorder(start_station_name, -number_of_rides), y = number_of_rides,
             fill = member_casual)) + 
  geom_col() +
  labs(
    title = "Top 10 Stations with Highest rides",
    x = "Station Name",
    y = "Number of Rides",
    caption = "Source: https://divvy-tripdata.s3.amazonaws.com/index.html",
    fill = "Member Type"
  ) + 
  geom_text(aes(label = number_of_rides),
            position = position_stack(vjust = 0.5)) +
  theme(axis.text.x = element_text(angle = 60, vjust = 0.5, size = 10),
        plot.title = element_text(size = 18))
```

### The Distribution of Rides by Quarter and Member Type.
``` {r graph of quarter}
all_trips_v2 |>
  group_by(Q, member_casual) |>
  summarise(number_of_rides = n())|>
  arrange(Q) |>
  mutate(Q = factor(
    Q,
    levels = c("Q2_2019", "Q3_2019", "Q4_2019", "Q1_2020"),
    ordered = TRUE)) |>
  ggplot(aes(x = Q, y = number_of_rides, fill = member_casual)) + 
  geom_col() + 
  labs(
    title = "The Number of Rides by Quarter.",
    subtitle = "From Q2-2019 to Q1-2020",
    x = "Quarter",
    y = "Number of Rides",
    caption = "Source: https://divvy-tripdata.s3.amazonaws.com/index.html",
    fill = "Member Type"
  ) +
  theme(plot.title = element_text(size = 18))
```


## **My Recommendations**

### 1. We should launch our marketing campaigns in the 2nd-Quarter to 3rd-Quarter for more reach our customers.

### 2. We can start a marketing pre-launch at 
+ **Streeter Dr & Grand Avw**, 
+ **Lake Shore Dr & Monroe St**, 
+ **Michigan Ave & Oak St **, **Lake Shore Dr & North Blvd **, 
+ **Lake Shore Dr & North Blvd **

### 3. The messages on canpaign should focus on the average duration of casual rider.